
```{python}
#| echo : false
#| warning: false
import arviz as az
import numpy as np
import xarray as xr
xr.set_options(display_expand_attrs=False)
az.style.use("arviz-variat")
```

```{r}
#| echo: false
#| warning: false
#| message: false
library(posterior)
library(ggplot2)
library(loo)
library(bayesplot)
library(priorsense)
theme_set(bayesplot::theme_default(base_family = "sans", base_size = 14))
library(tinytable)
```

## Sponsors

- Bayesian Workflow group at Aalto University
- EnvStat at University of Helsinki
- PyMC Labs

## Outline

- Tools and posterior draws objects (9:00-9:45)
- Convergence and uncertainty (9:50-10:55)
- Model evaluation and critique (11:00-12:00)

Work book: n-kall.github.io/stancon2026-workflow-tools

# Tools

## R

- [posterior](https://mc-stan.org/posterior/): Conversion, manipulation, and summarization of draws from posterior and prior distributions.
- [bayesplot](https://mc-stan.org/bayesplot/): Visual checks and summaries.
- [loo](https://mc-stan.org/loo/): Model comparison using leave-one-out cross-validation and related methods.
- [priorsense](https://n-kall.github.io/priorsense/): Prior diagnostics and sensitivity analysis.

## Python
- [arviz-base](https://python.arviz.org/projects/base/en/latest/): Data related functionality, including converters from different PPLs.
- [arviz-stats](https://python.arviz.org/projects/stats/en/latest/): Statistical functions and diagnostics.
- [arviz-plots](https://python.arviz.org/projects/plots/en/latest/): Visual checks and summaries built on top of arviz-stats and arviz-base.
- [arviz](https://python.arviz.org/en/latest/): a meta-package that imports all the above and provides a single namespace for users.

# Posterior objects

`posterior` provides objects for draws that mimic built-in types `matrix`,
`array` `data.frame`, `list`

`arviz` provides the `DataTree` object for posterior draws.

## Example

::: {.panel-tabset}

### Python

```{python}
#| echo: false
dt = az.load_arviz_data("centered_eight")
dt
```

### R

```{r}
#| echo: false

posterior::example_draws() |> posterior::as_draws_df()

```

:::

# MCMC diagnostics

## R-hat

- Convergence diagnostic
- Measure of ratio between within-chain and between-chain variance
- R-hat > 1.01 indicates lack of convergence

```{r}
sample <- readRDS("../models/prerun/sample.rds")
summarise_draws(sample, rhat) |> tt()
```

## Nested R-hat

- Convergence diagnostic when using many short chains
- Chains are grouped into superchains (sharing initialization)
- Classic R-hat may wrongly indicate lack of convergence, but nested R-hat does not

```{r}
#| echo: false
many_short_chains <- readRDS("../models/prerun/many_short_chains.rds")

summarise_draws(
    many_short_chains,
    rhat,
    nested_rhat = ~ rhat_nested(.x, superchain_ids = attr(many_short_chains, "superchain_ids"))
) |>
tt()
```


## Effective sample size

## Monte Carlo standard error

## Pareto-k

# Predictive checking

## Prior predictive checks

## Posterior predictive checks

# Model comparison

## Leave-one-out cross-validation

# Sensitivity checking

## Power-scaling sensitivity checks
