# Working with draws objects in R

```{r}
#| echo: false
#| warning: false
library(posterior)

```

The posterior package provides the `draws` object type. These are used for draws
generated from sampling. The benefit of these objects is that chains, iterations
and variables are all kept. The objects mirror common base R
objects: `draws_df`, `draws_list`, `draws_array`, `draws_matrix`. You can
transform from one type to another with `as_draws_*` functions. While the
package is called "posterior", the draws objects can contain draws from prior or
posterior.

Useful functions for summarising and manipulating draws are:

- `summarise_draws()`: generate a summary of the draws based on some functions
- `subset_draws()`: subset draws by variable, chain, iteration, draw
- `mutate_variables()`: create new variables, which can reference existing ones

`summarise_draws()` takes as arguments summary functions, such as `mean`, `median` or covergence diagnostics such as `rhat` or `ess_bulk`. For example:

```{r}
draws <- example_draws()
summarise_draws(draws, mean, sd, rhat)

```

For further details, see the [documentation for the
package](https://mc-stan.org/posterior).
