---
title: "Exercises"

---

::: {.content-visible when-profile="r"}

```{r}
library(posterior)
library(priorsense)
library(bayesplot)

```

:::


::: {.content-visible when-profile="python"}

```{python}
import arviz as az

```

:::


# Draws objects

For this set of exercises we will use draws from the classic eight schools model. These draws are included in both `posterior` and ArviZ.

::: {.content-visible when-profile="r"}

```{r}
eight_schools_draws <- example_draws("eight_schools")

```

:::

::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


## Understanding draws objects

Inspect the posterior object. How many chains, iterations and variables does it contain?

::: {.content-visible when-profile="r"}

You can print the draws object, or you can use the `nchains()`, `niterations()`
and `nvariables()` functions.

```{r}
# Your code here

```

:::

::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
nchains(eight_schools_draws)
niterations(eight_schools_draws)
nvariables(eight_schools_draws)

```

:::

::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::

## Subsetting draws

Extract only the first chain.

::: {.content-visible when-profile="r"}

You can use the `subset_draws()` function, specifying the `chain` argument.

```{r}
# Your code here

```

:::


::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  subset_draws(chain = 1)

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::



Extract only the first 10 iterations, from all the chains.

::: {.content-visible when-profile="r"}

You can use the `subset_draws()` function, specifying the `iteration` argument.

```{r}
# Your code here

```

:::

::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}


```{r}
eight_schools_draws |>
  subset_draws(iteration = 1:10)

```

:::

::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::


## Thinning draws

Thin the draws so that only half of the draws are included. Then try automatic thinning.

::: {.content-visible when-profile="r"}

You can use the function `thin_draws()`, specifying the `thin` argument.

```{r}
# Your code here

```

:::

::: {.content-visible when-profile="python"}

You can use the function ...

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}


```{r}
thin_draws(eight_schools_draws, thin = 2)

thin_draws(eight_schools_draws)

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::


##


## Summarising draws

Extract the variable `mu` and summarise it via mean and sd.

::: {.content-visible when-profile="r"}

You can use the functions `subset_draws()`, specifying the `variable` argument and `summarise_draws()`, specifying the functions.

```{r}
# Your code here

```

:::

::: {.content-visible when-profile="python"}

You can use the functions ...

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  subset_draws(variable = "mu") |>
  summarise_draws(mean, sd)

```

:::

::: {.content-visible when-profile="pythonsolutions"}


```{python}


```

:::


## Creating new variables

Create a new variable that is the difference between school 1 and school 2 means. Call it `diff_1_2`. Do the same with school 3 and school 4. Then summarise these new variables with median, 0.3 and 0.7 quantiles.

::: {.content-visible when-profile="r"}

You can use the functions `mutate_variables()` and `quantile2()`

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  mutate_variables(
    diff_1_2 = `theta[2]` - `theta[1]`,
    diff_3_4 = `theta[3]` - `theta[4]`
  ) |>
  subset_draws(
    variable = c("diff_1_2", "diff_3_4")
  ) |>
  summarise_draws(median, ~quantile2(.x, c(0.3, 0.7)))
```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::


# Convergence diagnostics

## R-hat

Calculate the R-hat for all the variables in the model. Which variables have high R-hat (> 1.01)?

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` and `rhat`

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  summarise_draws(rhat)

```

:::



::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::


## Effective sample size (ESS)

Calculate the bulk and tail ESS for all the variables in the model.

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` and `rhat()`

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  summarise_draws(ess_bulk, ess_tail)

```

:::



::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::


## Monte Carlo standard error

Calculate the mean of each variable, and also the Monte Carlo standard error of the mean.

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` and `mcse_mean()`

```{r}
# Your code here

```

:::


::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  summarise_draws(mean, mcse_mean)

```

:::


::: {.content-visible when-profile="python"}

```{python}


```

:::


Then do the same for the 0.05 and 0.95 quantiles. Think about how the Monte Carlo standard error might influence how you report the quantiles.

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` and `mcse_quantile()`

```{r}
# Your code here

```

:::


::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  summarise_draws(quantile2, mcse_quantile)

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::



## Pareto diagnostics

Calculate the minimum sample size for stable estimates for each variable in the model. Which has the highest minimum sample size?

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` and `pareto_min_ss()`

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}


```{r}
eight_schools_draws |>
  summarise_draws(pareto_min_ss)

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::

# Plotting posterior draws

## Marginal posteriors

Plot the marginal posteriors for the school means.

::: {.content-visible when-profile="r"}

You can use the functions `mcmc_intervals()` or `mcmc_areas()`.

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}


```{r}
eight_schools_draws |>
  subset_draws(variable = "theta") |>
  mcmc_intervals()

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::


## Pairs plot

Plot the variables `mu` and `tau` in an pairs plot.


::: {.content-visible when-profile="r"}

You can use the functions `mcmc_pairs()`

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}


```{r}
eight_schools_draws |>
    subset_draws(variable = c("mu", "tau")) |>
    mcmc_pairs()

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}


```

:::

# Prior predictive check

We can generate prior predictions from the eight schools model, using the following function.

::: {.content-visible when-profile="r"}

```{r}
eight_schools_prior <- function(ndraws, mu_prior_sd = 1, tau_prior_sd = 1, sigma_obs = c(15, 10, 16, 11, 9, 11, 10, 18)) {

  J <- length(sigma_obs)

  mu <- rnorm(ndraws, 0, mu_prior_sd)
  tau <- abs(rnorm(ndraws, 0, tau_prior_sd))

  theta <- sapply(1:J, \(j) rnorm(ndraws, mu, tau))
  yrep <- sapply(1:J, \(j) rnorm(ndraws, theta[, j], sigma_obs[j])) 
  
  prior_draws <- as_draws(yrep)
  variables(prior_draws) <- paste0("yrep[", 1:J, "]")
  
  return(prior_draws)
}
```

:::



::: {.content-visible when-profile="python"}

```{python}


```

:::


Generate 1000 prior predictive draws, and plot the distributions for each school.
Try with different `mu_prior_sd` and `tau_prior_sd` values (e.g. 1, 10, 100).

::: {.content-visible when-profile="r"}

You can use the `ppd_intervals` function.

```{r}
prior_predictive_draws <- eight_schools_prior(ndraws = 1000, mu_prior_sd = 1, tau_prior_sd = 1)
# Continue the code
```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_prior(ndraws = 1000, mu_prior_sd = 1, tau_prior_sd = 1) |>
  ppd_intervals()

eight_schools_prior(ndraws = 1000, mu_prior_sd = 0.01, tau_prior_sd = 1) |>
  ppd_intervals()

eight_schools_prior(ndraws = 1000, mu_prior_sd = 100, tau_prior_sd = 1) |>
  ppd_intervals()

eight_schools_prior(ndraws = 1000, mu_prior_sd = 1, tau_prior_sd = 0.01) |>
  ppd_intervals()
  
eight_schools_prior(ndraws = 1000, mu_prior_sd = 1, tau_prior_sd = 100) |>
  ppd_intervals()
```

:::



::: {.content-visible when-profile="pythonsolutions"}

```{python}
# Your code here

```

:::

## Posterior predictive checks

We can create posterior predictive draws from our posterior draws and plot against our actual observations.

::: {.content-visible when-profile="r"}
```{r}

sigma_obs <- c(15, 10, 16, 11, 9, 11, 10, 18)

eight_schools_post_pred <- eight_schools_draws |>
  mutate_variables(
    `yrep[1]` = rnorm(length(`theta[1]`), `theta[1]`, sigma_obs[1]),
    `yrep[2]` = rnorm(length(`theta[2]`), `theta[2]`, sigma_obs[2]),
    `yrep[3]` = rnorm(length(`theta[3]`), `theta[3]`, sigma_obs[3]),
    `yrep[4]` = rnorm(length(`theta[4]`), `theta[4]`, sigma_obs[4]),
    `yrep[5]` = rnorm(length(`theta[5]`), `theta[5]`, sigma_obs[5]),
    `yrep[6]` = rnorm(length(`theta[6]`), `theta[6]`, sigma_obs[6]),
    `yrep[7]` = rnorm(length(`theta[7]`), `theta[7]`, sigma_obs[7]),
    `yrep[8]` = rnorm(length(`theta[8]`), `theta[8]`, sigma_obs[8])
  ) |>
  subset_draws(variable = "yrep") |>
  as_draws_matrix()

# Or alternatively in a nicer way, using rvars

eight_schools_post_pred <- eight_schools_draws |>
  as_draws_rvars() |>
  mutate_variables(
    yrep = rvar_rng(
      rnorm,
      n = length(sigma_obs),
      mean = theta,
      sd = sigma_obs
    )
  ) |>
  subset_draws(variable = "yrep") |>
  as_draws_matrix()

```
:::

::: {.content-visible when-profile="python"}

:::

Plot the posterior predictions on top of the actual observations. Then use the PIT-ECDF plot.

::: {.content-visible when-profile="r"}
You can use the `ppc_intervals()` function.
```{r}
y <- c(28,  8, -3,  7, -1,  1, 18, 12)
# Your code here
```
:::

::: {.content-visible when-profile="rsolutions"}
```{r}
y <- c(28,  8, -3,  7, -1,  1, 18, 12)
ppc_intervals(y, eight_schools_post_pred)

ppc_pit_ecdf(y, eight_schools_post_pred, method = "correlated")
```
:::

::: {.content-visible when-profile="python"}
```{python}
# Your code here
```
:::

::: {.content-visible when-profile="pythonsolutions"}
```{r}

```
:::

