---
title: "Exercises"

---

::: {.content-visible when-format="org"}

```{=org}
#+PROPERTY: header-args:R :session *stancon-r* :results output :exports both
```

:::

::: {.content-visible when-profile="r"}
First load the environment and install the required packages:
```{r}
renv::restore(prompt = FALSE)

```

:::

::: {.content-visible when-profile="python"}

```{python}
#| eval: false
import shutil
import subprocess

if shutil.which("uv") is None:
    raise RuntimeError(
        "uv is not installed.\n"
        "Install it from https://docs.astral.sh/uv/getting-started/installation/\n"
        "Then restart the kernel and run this cell again."
    )

subprocess.run(["uv", "sync"], check=True)

print("Environment created successfully.")
print("Now change the kernel to uv: stancon2026-workflow-tools")
```

:::


::: {.content-visible when-profile="r"}

```{r}
library(posterior)
library(priorsense)
library(bayesplot)

```

:::


::: {.content-visible when-profile="python"}

```{python}
import arviz as az
```

:::


# Part 1: Posterior draws objects

For this set of exercises we will use draws from the classic eight schools model. These draws are included in both `posterior` and `ArviZ`.

The eight schools model is a meta-analysis model of standardized test results in different schools. The values `theta[1]` to `theta[8]` are the mean results for each school. `mu` is the population level mean, `tau` is the standard deviation of the population distribution.

::: {.content-visible when-profile="r"}

```{r}
eight_schools_draws <- example_draws("eight_schools")

```

:::

::: {.content-visible when-profile="python"}

```{python}
dt = az.load_arviz_data('non_centered_eight')
eight_schools_draws = dt.posterior
```

:::


## Understanding draws objects

Inspect the posterior object. How many chains, iterations and variables does it contain?

::: {.content-visible when-profile="r"}

You can print the draws object, or you can use the `nchains()`, `niterations()`
and `nvariables()` functions.

```{r}
# Your code here

```

:::

::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
nchains(eight_schools_draws)
niterations(eight_schools_draws)
nvariables(eight_schools_draws)

```

:::

::: {.content-visible when-profile="pythonsolutions"}

```{python}
eight_schools_draws.sizes["chain"]
eight_schools_draws.sizes["draw"]
len(eight_schools_draws.data_vars)
```

:::

## Subsetting draws

Extract only the first chain.

::: {.content-visible when-profile="r"}

You can use the `subset_draws()` function, specifying the `chain` argument.

```{r}
# Your code here

```

:::


::: {.content-visible when-profile="python"}

You can use the `sel()` method, specifying the `chain` argument.

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  subset_draws(chain = 1)

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}
eight_schools_draws.sel(chain=0)

```

:::



Extract only the first 10 iterations, from all the chains.

::: {.content-visible when-profile="r"}

You can use the `subset_draws()` function, specifying the `iteration` argument.

```{r}
# Your code here

```

:::

::: {.content-visible when-profile="python"}

You can use the `sel()` method, specifying the `draw` argument and using the `slice()` function.

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}


```{r}
eight_schools_draws |>
  subset_draws(iteration = 1:10)

```

:::

::: {.content-visible when-profile="pythonsolutions"}

```{python}
eight_schools_draws.sel(draw=slice(0, 10))

```

:::


## Thinning draws

Thin the draws so that only half of the draws are included. Then try automatic thinning.

::: {.content-visible when-profile="r"}

You can use the function `thin_draws()`, optionally specifying the `thin` argument.

```{r}
# Your code here

```

:::

::: {.content-visible when-profile="python"}

You can use the function `az.thin()`, optionally specifying the `factor` argument.

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}


```{r}
thin_draws(eight_schools_draws, thin = 2)

```

```{r}
thin_draws(eight_schools_draws)

```


:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}
az.thin(eight_schools_draws, factor=2)

```

```{python}
az.thin(eight_schools_draws)

```

:::


## Summarising draws

Extract the variable `mu` and summarise it via mean and sd.

::: {.content-visible when-profile="r"}

You can use the functions `subset_draws()`, specifying the `variable` argument and `summarise_draws()`, specifying the functions.

```{r}
# Your code here

```

:::

::: {.content-visible when-profile="python"}

You can use the functions `az.summary()`, specifying the `var_names` argument and the `kind` argument.

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  subset_draws(variable = "mu") |>
  summarise_draws(mean, sd)

```

:::

::: {.content-visible when-profile="pythonsolutions"}


```{python}
# XXX do we want to show this?
az.summary(eight_schools_draws, var_names=["mu"], kind="stats")

## or this?
az.summary(eight_schools_draws, var_names=["mu"], kind="stats")[["mean", "sd"]]

```

:::


## Creating new variables

Create a new variable that is the difference between school 1 (`theta[1]` and school 2 (`theta[2]`) means. Call it `diff_1_2`. Do the same with school 3 (`theta[3]`) and school 4 (`theta[4]`). Then summarise these new variables with median, 0.3 and 0.7 quantiles. 
::: {.content-visible when-profile="r"}

You can use the functions `mutate_variables()`, `summarise_draws()` and `quantile2()`. You can specify variable names that include `[]` with `` ` `` e.g. `` `theta[1]` ``.


```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  mutate_variables(
    diff_1_2 = `theta[2]` - `theta[1]`,
    diff_3_4 = `theta[3]` - `theta[4]`
  ) |>
  subset_draws(
    variable = c("diff_1_2", "diff_3_4")
  ) |>
  summarise_draws(median, ~quantile2(.x, c(0.3, 0.7)))
```

:::


::: {.content-visible when-profile="pythonsolutions"}

The simplest version is to call `az.summary()` with the `kind` argument set to `"stats_median"` and the `ci_prob` argument set to `0.40`.
```{python}
import xarray as xr

diff_1_2 = eight_schools_draws["theta"].isel(school=1) - eight_schools_draws["theta"].isel(school=0)
diff_3_4 = eight_schools_draws["theta"].isel(school=3) - eight_schools_draws["theta"].isel(school=2)

diffs = xr.Dataset({"diff_1_2": diff_1_2, "diff_3_4": diff_3_4})

az.summary(diffs, kind="stats_median", ci_prob=0.40)

```
But a more explicit version is to do.

```{python}
import pandas as pd
result = pd.DataFrame({
    "median": diffs.median(dim=["chain", "draw"]).to_pandas(),
    "q30": diffs.quantile(0.3, dim=["chain", "draw"]).to_pandas(),
    "q70": diffs.quantile(0.7, dim=["chain", "draw"]).to_pandas(),
})

result

```
Optionally you can use `.sel` and reference the school by name like `eight_schools_draws["theta"].sel(school="Choate")`

:::

## Marginal posteriors

Plot the marginal posteriors for the school means.

::: {.content-visible when-profile="r"}

You can use the functions `mcmc_intervals()` or `mcmc_areas()`.

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

You can use the function `az.plot_forest()`

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}


```{r}
eight_schools_draws |>
  subset_draws(variable = "theta") |>
  mcmc_intervals()

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}
az.plot_forest(eight_schools_draws, var_names=["theta"])
```

:::


## Pairs plot

Plot the variables `mu` and `tau` in an pairs plot.


::: {.content-visible when-profile="r"}

You can use the functions `mcmc_pairs()`

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}


```{r}
eight_schools_draws |>
    subset_draws(variable = c("mu", "tau")) |>
    mcmc_pairs()

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}
az.plot_pair(eight_schools_draws, var_names=["mu", "tau"])

```

:::


# Part 2: Convergence diagnostics and uncertainty

## R-hat

Calculate the R-hat for all the variables in the draws object (`eight_schools_draws`). Which variables have high R-hat (> 1.01)?

::: {.content-visible when-profile="r"}

You can use the function `summarise_draws()` with `rhat`

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

You can use the function `az.rhat()`


```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  summarise_draws(rhat)

```

:::



::: {.content-visible when-profile="pythonsolutions"}

```{python}
az.rhat(eight_schools_draws)
```

:::


## Effective sample size (ESS)

Calculate the bulk and tail ESS for all the variables in the model.

::: {.content-visible when-profile="r"}

You can use the function `summarise_draws()` with `ess_tail` and `ess_bulk`

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  summarise_draws(ess_bulk, ess_tail)

```

:::



::: {.content-visible when-profile="pythonsolutions"}

```{python}
az.summary(eight_schools_draws)[["ess_bulk", "ess_tail"]]

```
:::


## Monte Carlo standard error

Calculate the mean of each variable, and also the Monte Carlo standard error of the mean.

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` with `mean` and `mcse_mean`

```{r}
# Your code here

```

:::


::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  summarise_draws(mean, mcse_mean)

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}
az.summary(eight_schools_draws)[["mean", "mcse_mean"]]
```

:::


Then do the same for the 0.05 and 0.95 quantiles. Think about how the Monte Carlo standard error might influence how you report the quantiles.

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` with `quantile2` and `mcse_quantile`

```{r}
# Your code here

```

:::


::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_draws |>
  summarise_draws(quantile2, mcse_quantile)
```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}
mcse_q05 = az.mcse(eight_schools_draws, method="quantile", prob=0.05).ds.expand_dims(summary=["mcse_5%"])
mcse_q95 = az.mcse(eight_schools_draws, method="quantile", prob=0.95).ds.expand_dims(summary=["mcse_95%"])

eti_05 = (
    az.eti(eight_schools_draws, prob=0.90)
    .ds.rename({"ci_bound": "summary"})
    .assign_coords(summary=["eti90_lb", "eti90_ub"])
)

summary = xr.concat([mcse_q05, mcse_q95, eti_05], dim="summary")

az.dataset_to_dataframe(summary, sample_dims=["summary"]).T
```

:::



## Pareto diagnostics

Calculate the minimum sample size for stable estimates for each variable in the model. Which has the highest minimum sample size?

::: {.content-visible when-profile="r"}

You can use the function `summarise_draws()` with `pareto_min_ss`

```{r}
# Your code here

```

:::



::: {.content-visible when-profile="python"}

```{python}
# Your code here

```

:::


::: {.content-visible when-profile="rsolutions"}


```{r}
eight_schools_draws |>
  summarise_draws(pareto_min_ss)

```

:::


::: {.content-visible when-profile="pythonsolutions"}

```{python}
eight_schools_draws.azstats.pareto_min_ss()
```

:::


# Part 3: Model evaluation and critique

We can generate prior predictions from the eight schools model, using the following function.

::: {.content-visible when-profile="r"}

```{r}
eight_schools_prior <- function(ndraws, mu_prior_sd = 1, tau_prior_sd = 1, sigma_obs = c(15, 10, 16, 11, 9, 11, 10, 18)) {

  J <- length(sigma_obs)

  mu <- rnorm(ndraws, 0, mu_prior_sd)
  tau <- abs(rnorm(ndraws, 0, tau_prior_sd))

  theta <- sapply(1:J, \(j) rnorm(ndraws, mu, tau))
  yrep <- sapply(1:J, \(j) rnorm(ndraws, theta[, j], sigma_obs[j])) 
  
  prior_draws <- as_draws(yrep)
  variables(prior_draws) <- paste0("yrep[", 1:J, "]")
  
  return(prior_draws)
}
```

:::



::: {.content-visible when-profile="python"}

```{python}
import numpy as np
import pymc as pm
import pymc.dims as pmd

def make_eight_schools_model(mu_prior_sd=1, tau_prior_sd=1, observe=False):
  coords = {"school": eight_schools_draws.coords["school"].values}
  sigma_obs=xr.DataArray(np.array([15, 10, 16, 11, 9, 11, 10, 18]), coords=coords, dims=("school"))
  
  with pm.Model(coords=coords) as model:
    mu = pmd.Normal("mu", mu=0, sigma=mu_prior_sd)
    tau = pmd.HalfNormal("tau", sigma=tau_prior_sd)
    theta = pmd.Normal("theta", mu=mu, sigma=tau, dims=("school"))
    yrep = pmd.Normal("yrep", mu=theta, sigma=sigma_obs, dims=("school"), observed= dt["observed_data"]["obs"] if observe else None)
    
  return model

def eight_schools_prior(ndraws=1000, mu_prior_sd=1, tau_prior_sd=1):
  model = make_eight_schools_model(mu_prior_sd=mu_prior_sd, tau_prior_sd=tau_prior_sd)
  with model:
    prior_predictive_draws = pm.sample_prior_predictive(draws=ndraws)
  return prior_predictive_draws

```

:::


Generate 1000 prior predictive draws, and plot the distributions for each school.
Try with different `mu_prior_sd` and `tau_prior_sd` values (e.g. 1, 10, 100).

::: {.content-visible when-profile="r"}

You can use the `ppd_intervals` function.

```{r}
prior_predictive_draws <- eight_schools_prior(
  ndraws = 1000,
  mu_prior_sd = 1,
  tau_prior_sd = 1
)

# Your code here
```

:::



::: {.content-visible when-profile="python"}

```{python}
prior_predictive_draws = eight_schools_prior(
  ndraws=1000,
  mu_prior_sd=1,
  tau_prior_sd=1
)

# Your code here
```

:::

::: {.content-visible when-profile="rsolutions"}

```{r}
eight_schools_prior(ndraws = 1000, mu_prior_sd = 1, tau_prior_sd = 1) |>
  ppd_intervals()

eight_schools_prior(ndraws = 1000, mu_prior_sd = 0.01, tau_prior_sd = 1) |>
  ppd_intervals()

eight_schools_prior(ndraws = 1000, mu_prior_sd = 100, tau_prior_sd = 1) |>
  ppd_intervals()

eight_schools_prior(ndraws = 1000, mu_prior_sd = 1, tau_prior_sd = 0.01) |>
  ppd_intervals()
  
eight_schools_prior(ndraws = 1000, mu_prior_sd = 1, tau_prior_sd = 100) |>
  ppd_intervals()
```

:::



::: {.content-visible when-profile="pythonsolutions"}

```{python}
az.plot_forest(eight_schools_prior(ndraws=1000, mu_prior_sd=1, tau_prior_sd=1), group="prior", var_names=["yrep"])
az.plot_forest(eight_schools_prior(ndraws=1000, mu_prior_sd=0.01, tau_prior_sd=1), group="prior", var_names=["yrep"])
az.plot_forest(eight_schools_prior(ndraws=1000, mu_prior_sd=100, tau_prior_sd=1), group="prior", var_names=["yrep"])
az.plot_forest(eight_schools_prior(ndraws=1000, mu_prior_sd=1, tau_prior_sd=0.01), group="prior", var_names=["yrep"])
az.plot_forest(eight_schools_prior(ndraws=1000, mu_prior_sd=1, tau_prior_sd=100), group="prior", var_names=["yrep"])
```

:::

## Posterior predictive checks

We can create posterior predictive draws from our posterior draws and plot against our actual observations.

::: {.content-visible when-profile="r"}
```{r}

sigma_obs <- c(15, 10, 16, 11, 9, 11, 10, 18)

eight_schools_post_pred <- eight_schools_draws |>
  mutate_variables(
    `yrep[1]` = rnorm(length(`theta[1]`), `theta[1]`, sigma_obs[1]),
    `yrep[2]` = rnorm(length(`theta[2]`), `theta[2]`, sigma_obs[2]),
    `yrep[3]` = rnorm(length(`theta[3]`), `theta[3]`, sigma_obs[3]),
    `yrep[4]` = rnorm(length(`theta[4]`), `theta[4]`, sigma_obs[4]),
    `yrep[5]` = rnorm(length(`theta[5]`), `theta[5]`, sigma_obs[5]),
    `yrep[6]` = rnorm(length(`theta[6]`), `theta[6]`, sigma_obs[6]),
    `yrep[7]` = rnorm(length(`theta[7]`), `theta[7]`, sigma_obs[7]),
    `yrep[8]` = rnorm(length(`theta[8]`), `theta[8]`, sigma_obs[8])
  ) |>
  subset_draws(variable = "yrep") |>
  as_draws_matrix()

# Or alternatively in a nicer way, using rvars

eight_schools_post_pred <- eight_schools_draws |>
  as_draws_rvars() |>
  mutate_variables(
    yrep = rvar_rng(
      rnorm,
      n = length(sigma_obs),
      mean = theta,
      sd = sigma_obs
    )
  ) |>
  subset_draws(variable = "yrep") |>
  as_draws_matrix()

```
:::

::: {.content-visible when-profile="python"}

```{python}
with make_eight_schools_model() as model:
    posterior_predictive_draws = pm.sample_posterior_predictive(eight_schools_draws, var_names=["yrep"])

posterior_predictive_draws
```
:::

Plot the posterior predictions on top of the actual observations. Then use the PIT-ECDF plot.

::: {.content-visible when-profile="r"}
You can use the `ppc_intervals()` and `ppc_pit_ecdf()` functions.
```{r}
y <- c(28,  8, -3,  7, -1,  1, 18, 12)
# Your code here
```
:::

::: {.content-visible when-profile="rsolutions"}
```{r}
y <- c(28,  8, -3,  7, -1,  1, 18, 12)
ppc_intervals(y, eight_schools_post_pred)

ppc_pit_ecdf(y, eight_schools_post_pred, method = "correlated")
```
:::

::: {.content-visible when-profile="python"}
You can use the `az.plot_ppc_pit()` function.
```{python}
y = np.array([28,  8, -3,  7, -1,  1, 18, 12])
# Your code here
```
:::

::: {.content-visible when-profile="pythonsolutions"}
```{python}
y = np.array([28,  8, -3,  7, -1,  1, 18, 12])
posterior_predictive_draws["observed_data"]["yrep"] = xr.DataArray(y, dims=["school"])
az.plot_ppc_pit(posterior_predictive_draws, var_names=["yrep"])
```
:::


## Sensitivity checks

Check for prior and likelihood sensitivity in the eight schools model. First check by power-scaling all priors jointly, then select only the "mu" and only the "tau" prior separately.


::: {.content-visible when-profile="r"}
You can use `powerscale_sensitivity()` with the `prior_selection` argument.

```{r}
eight_schools_draws_ps <- example_powerscale_model("eight_schools")$draws
# Your code here
```
:::

::: {.content-visible when-profile="rsolutions"}
```{r}
eight_schools_draws_ps <- example_powerscale_model("eight_schools")$draws

powerscale_sensitivity(eight_schools_draws_ps, variable = c("mu", "tau", "theta"))
powerscale_sensitivity(eight_schools_draws_ps, variable = c("mu", "tau", "theta"), prior_selection = "mu")
powerscale_sensitivity(eight_schools_draws_ps, variable = c("mu", "tau", "theta"), prior_selection = "tau")

```
:::

::: {.content-visible when-profile="python"}
```{python}
eight_schools_draws
with make_eight_schools_model(observe=True) as model:
  pm.stats.compute_log_prior(
    dt,
    var_names=["mu", "tau", "theta"],
  #  extend_inferencedata=False
    )
# Your code here
```
:::

::: {.content-visible when-profile="pythonsolutions"}
```{python}

az.psense_summary(dt)

```
:::


Next plot sensitivity using as density plots. Plot only the mu and tau variables.

::: {.content-visible when-profile="r"}


```{r}
# Your code here
```
:::

::: {.content-visible when-profile="rsolutions"}
You can use `powerscale_plot_dens()`.
```{r}
powerscale_plot_dens(eight_schools_draws_ps, variable = c("mu", "tau"))
powerscale_plot_dens(eight_schools_draws_ps, variable = c("mu", "tau"), prior_selection = "mu")
powerscale_plot_dens(eight_schools_draws_ps, variable = c("mu", "tau"), prior_selection = "tau")
```
:::

::: {.content-visible when-profile="python"}
```{python}
# Your code here
```
:::

::: {.content-visible when-profile="pythonsolutions"}
```{python}

az.plot_psense_dist(dt, var_names=["mu", "tau"])

```
:::

# Part 4: Bringing it all together

We have provided four sets of posterior draws from hierarchical models of observed migratory bird counts recorded between 2000 and 2020 at the [Hanko Bird Observatory (Halias)](https://halias.fi/tutkimus-ja-aineisto/).

For species $j$,

$y \sim \mathrm{Poisson}(\lambda_j)$

or

$y \sim \mathrm{NegativeBinomial}(\lambda_j, \phi).$

The species-specific abundances are linked through a hierarchical prior,

$\log(\lambda_j) \sim \mathrm{Normal}(\mu, \sigma)$

where $\mu$ represents the average abundance across species and $\sigma$
controls the amount of pooling between species.

Your task is to explore the posterior draws and diagnostic outputs for the four
fitted models.

As you work through the diagnostics, try to identify which model corresponds to
each of the following situations:

- Convergence issues caused by poor chain mixing (for example, insufficient warmup).
- Inadequate fit to the data caused by the choice of observation model, shown by posterior predictive checks.
- Issues caused by priors in conflict with the likelihood.
- No major issues, although there is still room for model improvement.

Use posterior summaries, convergence diagnostics, posterior predictive checks,
and sensitivity analyses to guide your investigation.

You can also look at the Stan model (`birds_per_year.stan`) and consider how you might improve it.

::: {.content-visible when-profile="r"}


```{r}
# Use the following as a starting point

# Load the draws
draws1 <- readRDS("data/draws1.rds")
draws2 <- readRDS("data/draws2.rds")
draws3 <- readRDS("data/draws3.rds")
draws4 <- readRDS("data/draws4.rds")

# Load the data fed to Stan
stan_data <- readRDS("data/stan_data.rds")

# Get the actual observations and the corresponding species
y <- stan_data$count
species <- stan_data$species # useful for grouped plots

# Create yrep matrices for posterior predictive checks
yrep1 <-  as_draws_matrix(subset_draws(draws1, "y_rep"))
yrep2 <-  as_draws_matrix(subset_draws(draws2, "y_rep"))
yrep3 <-  as_draws_matrix(subset_draws(draws3, "y_rep"))
yrep4 <-  as_draws_matrix(subset_draws(draws4, "y_rep"))
```

:::

::: {.content-visible when-profile="rsolutions"}
```{r}

```
:::

::: {.content-visible when-profile="python"}
```{python}
# Your code here
```
:::

::: {.content-visible when-profile="pythonsolutions"}
```{python}


```
:::
