---
title: "Workshop Exercises"
jupyter: python3
---

::: {.content-visible when-profile="r"}

```{r}
library(posterior)
library(priorsense)
library(bayesplot)

```

:::


::: {.content-visible when-profile="python"}

```{python}
import arviz as az

```

:::


# Draws objects

For this set of exercises we will use draws from the classic eight schools model. These draws are included in both `posterior` and ArviZ.

## Understanding draws objects

Inspect the posterior object. How many chains, iterations and variables does it contain?

::: {.content-visible when-profile="r"}

```{r}


```

:::

::: {.content-visible when-profile="python"}

```{python}


```

:::

## Subsetting draws

Extract only the first chain.

::: {.content-visible when-profile="r"}

You can use the `subset_draws()` function, specifying the `chain` argument.

```{r}


```

:::

::: {.content-visible when-profile="python"}

```{python}


```

:::


Extract only the first 10 iterations, from all the chains.

::: {.content-visible when-profile="r"}

You can use the `subset_draws()` function, specifying the `iteration` argument.

```{r}


```

:::

::: {.content-visible when-profile="python"}

```{python}


```

:::


## Thinning draws

Thin the draws so that only half of the draws are included. Then try the automatic thinning.

::: {.content-visible when-profile="r"}

You can use the function `thin_draws()`, specifying the `thin` argument.

```{r}


```

:::



::: {.content-visible when-profile="python"}

You can use the function ...

```{python}


```

:::


## 


## Summarising draws

Extract the variable `mu` and summarise it via mean and sd.

::: {.content-visible when-profile="r"}

You can use the functions `subset_draws()`, specifying the `variable` argument and `summarise_draws()`, specifying the functions.

```{r}



```

:::

::: {.content-visible when-profile="python"}

You can use the functions ...

```{python}

```

:::

## Creating new variables

Create a new variable that is the difference between school 1 and school 2 means. Call it `diff_1_2`. Do the same with school 3 and school 4. Then summarise these new variables with median, 0.3 and 0.7 quantiles.

::: {.content-visible when-profile="r"}

You can use the functions `mutate_variables()` and `quantile2()`

```{r}


```

:::



::: {.content-visible when-profile="python"}

```{python}


```

:::


# Convergence diagnostics

## R-hat

Calculate the R-hat for all the variables in the model. Which variables have high R-hat (> 1.01)?

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` and `rhat`

```{r}


```

:::



::: {.content-visible when-profile="python"}

```{python}


```

:::

## Effective sample size (ESS)

Calculate the bulk and tail ESS for all the variables in the model.

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` and `rhat()`

```{r}


```

:::



::: {.content-visible when-profile="python"}

```{python}


```

:::


## Monte Carlo standard error

Calculate the mean of each variable, and also the Monte Carlo standard error of the mean. The do the same for the 0.05 and 0.95 quantiles.

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` and `mcse_mean()`

```{r}


```

:::


::: {.content-visible when-profile="python"}

```{python}


```

:::



## Pareto diagnostics

Calculate the minimum sample size for stable estimates for each variable in the model. Which has the highest minimum sample size?

::: {.content-visible when-profile="r"}

You can use the functions `summarise_draws()` and `pareto_min_ss()`

```{r}


```

:::



::: {.content-visible when-profile="python"}

```{python}


```

:::
